Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-11-14 inst/doc/nonparametric-bayes.Rmd
tweaking

Decision making under uncertainty is a ubiquitous challenge of natural 
resource management and conservation.

The sudden collapse of fisheries and other ecosystems is an increasingly
widespread phenomenon and a pressing concern for ecological management
and conservation.  Ecological dynamics are frequently complex and difficult to 
measure, making uncertainty in our understanding a prediction a persistent
challenge to effective management. 
Decision-theoretic approaches provide a framework to determine the best 
sequence of actions in face of uncertainty, but only when that uncertainty
can be meaningfully quantified [@Fischer2009].

@@

Uncertainty enters the decision-making process at many levels: intrinsic 
stochasticity in biological processes, measurements, and implementation of 
policy [_e.g._ @Reed1979; @Clark1986; @Roughgarden1997; @Sethi2005], parameteric 
uncertainty [_e.g._ @Ludwig1982; @Hilborn1997; @McAllister1998; @Schapaugh2013],
and model or structural uncertainty [_e.g._ @Williams2001; @Cressie2009;  @Athanassoglou2012].
Of these, structural uncertainty incorporates the least a priori knowledge or 
assumptions and is generally the hardest to quantify. Typical approaches assume 
a weak notion of uncertainty where a correct or reasonable approximation of the dynamics
must be identified from among a handful of alternative models.  Here we consider an approach
that addresses uncertainty at each of these levels without assuming the dynamics follow
a particular (i.e. parametric) structure.

@@

@@

<!-- Integrate this into the paragraph, rather than as a footnote?? Make more consise? -->

[^1]: The concept of adaptive probing is one area that has explicitly
addressed this kind of uncertainty, reaching rather opposite conclusions
than what we observe here. Adaptive probing strategies follow from "Dual
Control" or "Active Adaptive Management" approaches (e.g. @Ludwig1982)
that can trade off short term utility by choosing actions that can reduce
uncertainty.  Adaptive probing strategies arise when it is valuable to
intentionally force a system far from the observed values even when the
expected value such actions is low, as it provides much faster learning
and consequent reduction of model uncertainty that can allow greater value
to be derived later on.  For instance, @Ludwig1982 show that it may be
advantageous to fish an unexploited population very heavily at first
to obtain a better estimate of the recruitment rate.  This intuitive
strategy when a population is governed by a Ricker or Beverton-Holt-like
dynamic would clearly be disastrous if instead the dynamics contained
an unforeseen tipping point.  The best way to learn where the edge lies
may be to walk up to it, but it is also the most dangerous.

@@

_Map_

@@

<!-- More Consise versions are now above 
- Parametric uncertainty

@@

- Structural uncertainty

@@

- Limits of state space

@@

-  _Note on "not magic"_: honest uncertainty + SDP

@@

<!--move to  much later -->

@@

_Do we need this?_

@@

- Background on non-parametric modeling.

@@

- Definition

- Previous application

@@

- Why it is particularly suited to these two problems
- (Why this is a novel application thereof)

@@

### Summary of approach

@@

- Underlying model 
- Available data 
- Value function

@@

- Statement of the models

@@

- Bayesian inference of parametric models

@@

- SDP via parametric models

@@

- Statement of model

... more on GP ... [Munch 2005]()

We also estimate a simple Gaussian Process defined by
a radial basis function kernel of two parameters: $\ell$, which gives
the characteristic length-scale over which correlation between two 
points (e.g. any two points $X_t, X_{t+1}$, and $X_{t+\tau}, X_{t+1+\tau}$)
in state-space decays, and $\sigma$, which gives the scale of the 
process noise by which observations $Y_{t+1}$ may differ from their
predicted values $X_{t+1}$ given an observation of the previous state,
$X_t$.

- Inference of the model

Also unlike parametric models, this posterior distribution is still
conditional on the training data. As such, the uncertainty near the
observed data.

@@

- SDP via the model

@@

- All models are "good fits" to the originally observed data. 
- (Simple model choice immediately leads us astray)
- Understanding the role of the prior

<!--

Though simple mechanistically motivated models offer the greatest potential 
to increase our basic understanding of ecological processes [@Cuddington2013; @Geritz2012], 
such models can be not only inaccurate but misleading when relied upon in a
quantitative decision making framework.  
-->

@@

These aspects are common to many conservation decision making problems, which thus merit
greater use of non-parametric approaches that can best take advantage of them.

@@

In our treatment here we have ignored the possibility of learning during the 
management phase, in which the additional observations of the stock size could
potentially improve parameter estimates.  While we intend to address this 
possibility in future work in the context of these non-parametric models,
we have not addressed it here for pedagogical reasons. In the context presented
here, it is clear that the differences in performance arise from differences
in the uncertainty inherent in the model formulations, rather than from 
differing abilities to learn.  Because we consider a threshold system, 
online learning would not change this generic feature of a lack of data in a
certain range of the state space which is better captured by the Gaussian process.

>>>>

Decision making under uncertainty is a ubiquitous challenge of natural 
resource management and conservation.  Ecological dynamics are
frequently complex and difficult to measure, making uncertainty in our
understanding a prediction a persistent challenge to effective management.
Decision-theoretic approaches provide a framework to determine the
best sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified [@Fischer2009].  The sudden
collapse of fisheries and other ecosystems has increasingly emphasized
the difficulties of formulating even qualitatively correct models of
the underlying processes.

@@

Uncertainty enters the decision-making process at many levels:
intrinsic stochasticity in biological processes, measurements, and
implementation of policy [_e.g._ @Reed1979; @Clark1986; @Roughgarden1997;
@Sethi2005], parameteric uncertainty [_e.g._ @Ludwig1982; @Hilborn1997;
@McAllister1998; @Schapaugh2013], and model or structural uncertainty
[_e.g._ @Williams2001; @Cressie2009;  @Athanassoglou2012].  Of these,
structural uncertainty incorporates the least a priori knowledge or
assumptions and is generally the hardest to quantify. Typical approaches
assume a weak notion of model uncertainty in which the correct model
(or reasonable approximation) of the dynamics must be identified from
among a handful of alternative models.  Here we consider an approach
that addresses uncertainty at each of these levels without assuming the
dynamics follow a particular (i.e. parametric) structure.

@@

Model uncertainty is particularly insidious when model predictions
must be made outside of the range of data on which the model was estimated.

@@

@@

<!--
_Map_
-->

@@

<!-- More consise versions are now above 
#### Parametric uncertainty

@@

#### Structural uncertainty

@@

#### Limits of state space

@@

<!--
-  _Note on "not magic"_: honest uncertainty + SDP
-->

@@

<!--move to  much later *Or do we need this at all?* -->

@@

@@

<!-- 
- Background on non-parametric modeling.  
-->

@@

<!--
- Definition
- Previous application
-->

@@

<!--
- Why it is particularly suited to these two problems
- (Why this is a novel application thereof)
-->

@@

<!-- skip this oveview? 
### Summary of approach
-->

@@

<!-- FIXME -->

- Underlying model 
- Available data

<!--
- Value function
-->

@@

<!-- 
- Statement of the models
-->

@@

<!-- Bayesian inference of parametric models -->

@@

<!-- SDP via parametric models -->

@@

<!-- Statement of model -->

<!-- FIXME ... say much more about the GP, as per Munch 2005a   -->
<!-- See much longer discusion above on "Background on GP".  Move some of that to here?  -->

We also estimate a simple Gaussian Process defined by
a radial basis function kernel of two parameters: $\ell$, which gives
the characteristic length-scale over which correlation between two 
points 
in state-space decays, and $\sigma$, which gives the scale of the 
process noise by which observations $Y_{t+1}$ may differ from their
predicted values $X_{t+1}$ given an observation of the previous state,
$X_t$. @Munch2005a gives an accessible introduction to the use of 
Gaussian Processes in providing a Bayesian nonparametric description
of the stock-recruitment relationship.

<!-- Inference of the BNP model -->

In contrast to the parametric models, this posterior distribution is still
conditional on the training data. As such, the uncertainty near the
observed data.

@@

<!-- SDP via the model -->

@@

In any modeling effort, models must be chosen for the task at hand.  
Though simple mechanistically motivated models offer the greatest potential 
to increase our basic understanding of ecological processes [@Cuddington2013; @Geritz2012], 
such models can be not only inaccurate but misleading when relied upon in a
quantitative decision making framework.  
In this paper we have tackled two aspects of uncertainty that are both common
to many ecological desicion-making problems and fundamentally challenging to
existing approaches which largely rely on parametric models:

@@

We have illustrated how the use of non-parametric approaches can provide
more reliable solutions in the sequential decision-making problem.

@@

In our treatment here we have ignored the possibility of learning during the 
management phase, in which the additional observations of the stock size could
potentially improve parameter estimates.  Of particular interest in the 
context of the extreme uncertainty considered here is the notion of 
"active adaptive management" or "adaptive probing", which may actively
seek to reduce uncertainty.  
The concept of adaptive probing is one area that has explicitly
addressed the extrapolation uncertainty addressed here, though 
typically without the additional issue of model uncertainty.  As a 
result, adaptive probing strategies suggest rather opposite conclusions
than what we observe here. Such adaptive probing or Dual
Control (e.g. @Ludwig1982) approaches 
trade off short term utility by choosing actions that can reduce
uncertainty.  Adaptive probing strategies arise when it is valuable to
intentionally force a system far from the observed values even when the
expected value such actions is low, as it provides much faster learning
and consequent reduction of model uncertainty that can allow greater value
to be derived later on.  For instance, @Ludwig1982 show that it may be
advantageous to fish an unexploited population very heavily at first
to obtain a better estimate of the recruitment rate.  This intuitive
strategy when a population is governed by a Ricker or Beverton-Holt-like
dynamic would clearly be disastrous if instead the dynamics contained
an unforeseen tipping point.  The best way to learn where the edge lies
may be to walk up to it, but it is also the most dangerous.  Future work
should attempt to understand when such active adaptive learning is valuable,
and when it will increase the risk of an irreversible transition.
